The role of habitat quality in host metapopulations exposed to environmentally persistent pathogens

### 2015-06-09 Manuscript/manuscript.tex
Title and abstract

\title{Modelling the creation of ecological traps by environmentally persistent pathogens}

@@

\begin{abstract} 
The distribution of habitat quality is expected to play an important role in the dynamics of wildlife metapopulations, with the expectation that high quality habitat serves to stabilize and maintain occupancy.  However, we might expect that the costs and benefits of different quality habitat might be modulated, or possibly even reversed, by the presence of an environmentally persistent pathogen.  In such cases, the presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk.  We explore this possibility through the development of a stochastic patch occupancy model in which we can vary both the environmental longevity of the pathogen, and the distribution of habitat quality in the host metapopulation.  This suggests that for sufficiently long-lived environmental reservoirs, high quality habitat can serve as an ecological trap, and can even be a net detriment to the host metapopulation relative to low quality habitat.  This inversion of the relative roles of high and low quality habitat highlights the importance of considering the interaction between spatial structure and pathogen transmission when managing wildlife populations exposed to an environmentally persistent pathogen.
\end{abstract}

@@

To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).  To isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).  Here again we performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by infectious hosts (as a measure of disease risk) and the final total effective host population size.

>>>>

\title{The role of habitat quality in host metapopulations exposed to environmentally persistent pathogens}

@@

\begin{abstract} 
The distribution of habitat quality is expected to play an important role in the dynamics of wildlife metapopulations, with the expectation that high quality habitat serves to stabilize and maintain occupancy.  However, we might expect that the costs and benefits of different quality habitat might be modulated, or possibly even reversed, by the presence of an environmentally persistent pathogen.  In such cases, the presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk.  We explore this possibility through the development of a stochastic patch occupancy model in which we can vary both the traits of the pathogen (virulence, transmission, environmental longevity) and the distribution of habitat quality in the host metapopulation.  This model suggests that for certain pathogens, high quality habitat can serve as an ecological trap, and can even be a net detriment to the host metapopulation relative to low quality habitat.  This inversion of the relative roles of high and low quality habitat highlights the importance of considering the interaction between spatial structure and pathogen transmission when managing wildlife populations exposed to an environmentally persistent pathogen.
\end{abstract}

@@

To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).  To isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).  Here again we performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by infectious hosts (as a measure of disease risk and the corresponding fitness costs) and the final total effective host population size.
